Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-07-05 paper_jasa.tex
conventional inference is chosen and all other terms are replaced

\begin {abstract}  
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. Until recently there were no formal visual methods in place for determining statistical significance of findings. This changed, when \citet{buja:2009} conceptually introduced two protocols for formal tests of visual findings. In this paper we take this a step further by comparing the lineup protocol \citep{buja:2009} against classical statistical testing  of the significance of regression model parameters. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol provides results comparable to the conventional tests, out-performs them when data is contaminated, and, there may be some super-individuals who yield better power than the conventional test even in the most difficult trials.
\end {abstract}

@@

However, measuring the strength of patterns seen in plots, and differences in individual perceptual ability, is something that is difficult and perhaps handicaps graphics  use among statisticians, where measuring probabilities is of primary importance. This has also been addressed in recent research. \citet{buja:2009} proposes a protocol that allow the testing of discoveries made from statistical graphics. This work represents a major advance for graphics, because it bridges the gulf between classical statistical inference procedures and exploratory data analysis. One of the protocols, the lineup, places the actual data plot among  a page of plots of null data, and asks a human judge to pick the plot that is different. Figure \ref{fig:test_category} shows an example lineup. Which plot do you think is the most different from the others? (The position of the actual data plot is provided in Section \ref{sec:category}.) Wrapped in a process that mirrors classical inference, where there is an explicit, a priori, null hypothesis, picking the plot of the data from the null plots represents a rejection of that null hypothesis. The null hypothesis typically derives from the task at hand, or the type of plot being made. The alternative encompasses all possible antitheses, all types of patterns that might be detected in the actual data plot, accounting for all possible deviations from the null without the requirement to specify these ahead of time. The probability of rejection can be quantified, along with Type I, and Type II error, and $p$-value and power can be defined and estimated.

@@

The protocol has only been informally tested until now. In the work described in this paper, the lineup protocol is compared head to head with the equivalent conventional test. Specifically, the lineup is examined in the context of a linear model setting, where we are determining the importance of including a variable in the model. This is not the envisioned environment for the use of the lineup -- actually it is likely the worst case scenario for visual inference. The intended use of lineups is where there is no existing test, and unlikely to be ever any numerical test. The thought is though, that the classical setting provides a benchmark for how well the lineup protocol works under controlled conditions, and will provide some assurance that they will work in scenarios where there is no benchmark. Testing is done using a human-subjects experiment using Amazon's Mechanical Turk \citep{turk}, using simulation to provide controlled conditions for assessing lineups. The results are compared with those of the conventional test.

@@

%\citet{buja:2009}, following from \cite{gelman:2004}, proposed two protocols that allow the testing of discoveries made from statistical graphics. This work represents a major advance for graphics, because it bridges the gulf between classical statistical inference procedures and exploratory data analysis.

%In this paper we compare the lineup protocol to classical statistical inference, specifically applied to the regression framework. We also present results of a human-subject study assessing the performance of individuals on lineup plots \citep{buja:2009} for testing significance of regression parameters.

@@

Table \ref{tbl:compare} illustrates the lineup protocol in relation to traditional hypothesis testing. Both start from the same place, the same null hypothesis. The conventional test statistic is the $t$-statistic, where the parameter estimate is divided by its standard error. In the lineup protocol, the test statistic is a plot of the data, here side-by-side boxplots are used, because the variable of interest is categorical and takes just two values. In traditional hypothesis testing the value of the test statistic is compared with all possible values of the sampling distribution, the distribution of the statistic if the null hypothesis is true. If it is extreme on this scale then the null hypothesis is rejected. In contrast in visual inference, the plot of the data is compared with plots of a finite number of samples drawn from the null distribution. If the actual data plot is selected as the most different, then this results in rejection of the null hypothesis.

\begin{table*}[hbtp]
\caption{Comparison of visual inference with existing inference technique.  }
\centering 
\begin{tabular}{llll} 
\hline
  & Traditional Inference &  Lineup Protocol \\ %[0.5ex] % inserts table %heading 
\hline
  Hypothesis & $H_0: \beta=0$ vs $H_1: \beta > 0$& $H_0: \beta=0$ vs $H_1: \beta > 0$\\
 & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} \\

@@

%This section outlines the concepts of visual inference in comparison to the procedures of classical statistical inference. Table \ref{tbl:compare} (derived from \citet{buja:2009}) gives a summarized overview of this comparison.

In general, we define $\theta$ to be a population parameter of interest, with $\theta \in \Theta$, the parameter space. Any null hypothesis $H_0$ then partitions the parameter space into $\Theta_0$ and $\Theta_0^c$, with $H_0: \theta \in \Theta_0$ versus $H_1: \theta \in \Theta_0^c$. A test statistic, $T(y)$, is a function that maps the sample into a numerical summary, that can be used to test the null hypothesis. The hypothesis test maps the test statistic into \{0, 1\}, based on whether $T(y)$ falls into the acceptance region, or the rejection region, respectively. $T(y)$ is assessed relative to null values of this statistic $T(y_0)$ which would be possible values of $T$ if $\theta \in \Theta$.  %Similarly, for any test statistic, we have a rejection region $R$ and, complementary to that, $R^c$,  that partitions the space of the test statistic, which is usually a subspace of the real numbers. 
For visual inference, unlike in the classical hypothesis test, the statistic is not a single value, but a graphical representation of the data chosen to display the strength of the parameter of interest, $\theta$. When the alternative hypothesis is true, it is expected that the plot of the actual data, the test statistic, will have visible feature(s) consistent with $\theta \in \Theta_0^c$, and that visual artifacts will not distinguish the test statistic as different when $H_1$ is not true. We will call a plot with this property a {\it visual statistic} for $\theta$. More formally,

@@

Power is an important consideration in deciding which test to use for solving a problem. %In classical testing, new test procedures are examined for their power under a variety of scenarios. 
For visual inference, this corresponds to the choice of plot to use. Analysts typically have a choice of plots to make, and a myriad of possible options such as reference grids, for any particular purpose. This is akin to different choices of statistics in classical hypothesis testing, for example, mean, median, or trimmed mean. One is typically better than another. For two different visual test statistics of the same actual data, one is considered to be better, if $T(y)$ is more easily distinguishable to the observer. Power is typically used to measure this characteristic of a test. %Power is the probability of rejecting $H_0$ when it is not true.  We can assess power therefore both empirically based on experimental data and through theoretical considerations.

@@

The following properties are only derived for the situation of a  fully parameterized  simulation study. This allows us to directly compare lineup tests against their classical counterparts and also identify  properties relevant for a quality assessment of  lineups, which are typically used  in non-parametric situations.

@@

%Now suppose $F_{|t|}(.;\delta)$ denotes the distribution function of an absolute value of $t$, the classical test statistic.
%, with non-centrality parameter $\delta$. 
Then the distribution function of the $p$-value $p_D$ under $H_0$ is uniform, since:  
\begin{eqnarray}\label{dist_p}
F{p_D}(p) &=& Pr(p_D \le p)=1-Pr(1-p_D \le 1-p) \nonumber \\
  &=& 1-Pr\left(F_{|t|}^{-1}(1-p_D) \le F_{|t|}^{-1}(1-p) \right) \nonumber \\
  &=& 1-Pr\left(|t_{obs}| \le F_{|t|}^{-1}(1-p)\right) \nonumber \\
  &=&%\left\{ \begin{array}{ll}
          1-F_{|t|}\left( F_{|t|}^{-1}(1-p)\right)=p \mbox{ ; under $H_0$} 
 %         1-F_{|t|}( F_{|t|}^{-1}(1-p); \delta) &\mbox{ ; under $H_1$} 
%       \end{array} \right.     
\end{eqnarray}

@@

In this setting, there are many established graphics that are used to evaluate and diagnose the fit of a regression model (e.g. \citet{cook:99}). Table \ref{tbl:stat_multiple} lists several common hypotheses related to the regression setting, and the visual test statistic that might be used for testing these hypotheses. These plots should be familiar, because they are all in common use. For example, to examine the effect of variable $X_j$ on $Y$, we would plot $Y$ against $X_j$ (cases 1-4). To assess whether the linear assumption is appropriate we would make a plot of residuals against fitted values (case 5). For the purposes of comparing visual against classical inference, we focus on cases 2 and 3, with a continuous and catgorical explanatory variable respectively.

@@

% Di: This is specific for the regression model
%Consider the hypothesis test $H_0: \beta_k=0$ against $H_1: \beta_k \ne 0$ in model \ref{multi}.
%%Now consider estimating the power of the visual test. 
%%Suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$. For the regression slope parameter $\beta$ suppose the associated test statistic in classical test be $t_{obs}$ with $p$-value $p_D$. By definition we have 
%%
%%$$p_D=Pr(|t| \ge t_{obs}| H_0)=1-F_{|t|}(t_{obs}) \Rightarrow |t_{obs}|=F_{|t|}^{-1}(1-p_D)$$
%%
%%\red{HH: the following paragraph should probably move up, but I'm not sure how to change it appropriately. Mahbub, could you take care of that?}
%%Now suppose $F_{|t|}(.;\delta)$ denotes the distribution function of an absolute value of $t$ distribution with non-centrality parameter $\delta$. Thus the distribution function of $p-value$ be  
%%\begin{eqnarray}\label{dist_p}
%%F{p_D}(p) &=& Pr(p_D \le p)=1-Pr(1-p_D \le 1-p) \nonumber \\
%%  &=& 1-Pr(F_{|t|}^{-1}(1-p_D) \le F_{|t|}^{-1}(1-p) ) \nonumber \\
%%  &=& 1-Pr(|t_{obs}| \le F_{|t|}^{-1}(1-p)) \nonumber \\
%%  &=&\left\{ \begin{array}{ll}
%%          1-F_{|t|}( F_{|t|}^{-1}(1-p))=p &\mbox{ ; under $H_0$} \\
%%          1-F_{|t|}( F_{|t|}^{-1}(1-p); \delta) &\mbox{ ; under $H_1$} 
%%       \end{array} \right.     
%%\end{eqnarray}
%%
%%Thus the density of $p_D$ is Uniform(0,1)  under $H_0$. As noted by \cite{Ruppert:2007}, under $H_1$ the density $$f_{p_D}(p_D; \delta)= \frac{f_{|t|}(F_{|t|}^{-1}(1-p_D);\delta)}{f_{|t|}(F_{|t|}^{-1}(1-p_D))}$$ derived from equation \eqref{dist_p}  is a right skewed distribution.  
%In a lineup plot we simulate $m-1$ residual data sets from null model where each of these $m-1$ null data sets produces a corresponding $p$-value $p_{0,i}$ and $p_{0,i} \sim \text{Uniform}(0,1)$ for $i = 1, ..., m-1$. Suppose  $p_0=\min_i(p_{0,i})$. Thus $p_0 \sim \text{Beta}(1,m-1)$.

The conventional test in this scenario uses $T= \hat{\beta_k}/ se(\hat{\beta_k})$ and rejects the null hypothesis if $T$ is extreme on the scale of a $t$ distribution with $n-p$ degrees of freedom.  It forms the benchmark upon which we evaluate the visual test. To calculate what we might expect for the power of the visual test, under perfect conditions, first assume that the observer is able to pick the plot with the smallest $p$-value from a lineup plot.  This leads to the decision to reject $H_0$ when $p_{D} < p_0$, where $p_{D}$ is the classical $p$-value as details given in Lemma \ref{lemma}. Thus the expected power of the visual test in this scenario is

@@

Three experiments were conducted to evaluate the effectiveness of the lineup protocol relative to the equivalent test statistic used in the regression scenario. The first two are classical settings where we would not expect the lineup protocol to do better than the traditional test, but we hope it to perform favorably. The third experiment is a scenario where assumptions required for the traditional test are violated, and we would expect the lineup protocol to outperform the traditional test. (Data and lineups used in the experiments are available in the supplementary material.)

@@

%\green{Did you in fact bin like this???}\red{For each parameter combination a set of 1000 actual data are generated and their associated p-values were computed. This produces an empirical distribution of p-value under alternative hypothesis. Finally, three data sets that have 16th, 50th and 84th quantile of the p-values were taken. Similar procedure was used to generate 19 null plots based on the empirical distribution of p-value under null hypothesis}. For added control, to ensure a signal in the simulated actual data a blocking structure was used to filter data sets. A 1000 sets were generated for each parameter combination and the traditional $t$-statistic and $p$-value associated with $H_0: \beta_2=0$ were calculated. The 3 replicates were drawn from each of three blocks of $p$-values: (0.0-$q_{33}$), ($q_{33}$-$q_{66}$), ($q_{66}$-1) where $q_i$ is the $i$th percentile in the distribution of the $p$-values. Additional control was applied to the 19 null plots. Because the distribution of these $p$-values should follow a Uniform(0,1) distribution, data sets were binned on this range by $p$-value, and a data set was randomly selected from each bin.

@@

In the real setting, where visual inference is to be useful, there will be no conventional tests or traditional $p$-values. Assessing the strength of perceived structure is a critical component of visual inference. In experiments 1 and 2, there is a $p$-value associated with the actual data plot in each lineup. As the $p$-value increases the proportion of correct responses falls (Figure \ref{fig:pval_pcorrect}), which is evidence of direct association between proportion of correct responses and conventional test $p$-values. When $p$-value is larger than 0.15, it is very uncommon for subjects to correctly identify the actual data plot in the lineup.

@@

%Note that $\hat{p}_D$ is bounded within $(0, 1/(m-1))$. This estimate has to be interpreted carefully, since the choice of a large lineup size, $m$, will make the value look `significant' in a traditional sense. Values close to the upper bound of $1/(m-1)$ should therefore be interpreted as `at least as high', rather than their actual numeric value.
%\hh{For $p$-values below $1/(m-1)$ visual $p$-values are very similar to actual $p$-values, the correlation for those values is: 0.36, 0.79, and -0.48 based on $n_1 = 44, n_2 = 45$, and $n_3= 26$ lineups, respectively.  }  \blue{This is not happening as we see in figure \ref{fig:pval_plot_signal}. But it is true for experiment 2, right?. }
%is equivalent to classical test $p$-value when $p$-value is less than or equal to $1/(m-1)$.  
%and beyond that the plot signal strength levels off on a point bigger than 0.05. Thus if plot signal strength is less than 0.05, we may decide to reject the null hypothesis.

@@

The purpose of this paper has been to examine the effectiveness of visual inference methods in direct comparison to existing inference methods. We need to be clear that this is not the purpose of visual inference generally: visual methods should not be seen as competitors to traditional inference.  The purpose here, is to  establish properties and  efficacy of visual testing procedures in order to use them in situations where traditional tests cannot be used. For this experiment the effect of $\beta_2$ was examined using side-by-side boxplots. Future experiments will be conducted to compare other regression parameters as described in Table \ref{tbl:stat_multiple} and assess sensitivity of power to modeling conditions.

@@

\begin {abstract} 
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. Until recently there were no formal visual methods in place for determining statistical significance of findings. This changed, when Buja et al.(2009) conceptually introduced two protocols for formal tests of visual findings. In this paper we take this a step further by comparing the lineup protocol (Buja et al.2009) against classical statistical testing  of the significance of regression model parameters. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol provides results equivalent to the conventional test and for some scenarios yields better power than the conventional test.
\end {abstract}

>>>>

\begin {abstract}  
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. Until recently there were no formal visual methods in place for determining statistical significance of findings. This changed, when \citet{buja:2009} conceptually introduced two protocols for formal tests of visual findings. In this paper we take this a step further by comparing the lineup protocol \citep{buja:2009} against conventional statistical testing  of the significance of regression model parameters. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol provides results comparable to the conventional tests, out-performs them when data is contaminated, and, there may be some super-individuals who yield better power than the conventional test even in the most difficult trials.
\end {abstract}

@@

However, measuring the strength of patterns seen in plots, and differences in individual perceptual ability, is something that is difficult and perhaps handicaps graphics  use among statisticians, where measuring probabilities is of primary importance. This has also been addressed in recent research. \citet{buja:2009} proposes a protocol that allow the testing of discoveries made from statistical graphics. This work represents a major advance for graphics, because it bridges the gulf between conventional statistical inference procedures and exploratory data analysis. One of the protocols, the lineup, places the actual data plot among  a page of plots of null data, and asks a human judge to pick the plot that is different. Figure \ref{fig:test_category} shows an example lineup. Which plot do you think is the most different from the others? (The position of the actual data plot is provided in Section \ref{sec:category}.) Wrapped in a process that mirrors conventional inference, where there is an explicit, a priori, null hypothesis, picking the plot of the data from the null plots represents a rejection of that null hypothesis. The null hypothesis typically derives from the task at hand, or the type of plot being made. The alternative encompasses all possible antitheses, all types of patterns that might be detected in the actual data plot, accounting for all possible deviations from the null without the requirement to specify these ahead of time. The probability of rejection can be quantified, along with Type I, and Type II error, and $p$-value and power can be defined and estimated.

@@

The protocol has only been informally tested until now. In the work described in this paper, the lineup protocol is compared head to head with the equivalent conventional test. Specifically, the lineup is examined in the context of a linear model setting, where we are determining the importance of including a variable in the model. This is not the envisioned environment for the use of the lineup -- actually it is likely the worst case scenario for visual inference. The intended use of lineups is where there is no existing test, and unlikely to be ever any numerical test. The thought is though, that the conventional setting provides a benchmark for how well the lineup protocol works under controlled conditions, and will provide some assurance that they will work in scenarios where there is no benchmark. Testing is done using a human-subjects experiment using Amazon's Mechanical Turk \citep{turk}, using simulation to provide controlled conditions for assessing lineups. The results are compared with those of the conventional test.

@@

%\citet{buja:2009}, following from \cite{gelman:2004}, proposed two protocols that allow the testing of discoveries made from statistical graphics. This work represents a major advance for graphics, because it bridges the gulf between conventional statistical inference procedures and exploratory data analysis.

%In this paper we compare the lineup protocol to conventional statistical inference, specifically applied to the regression framework. We also present results of a human-subject study assessing the performance of individuals on lineup plots \citep{buja:2009} for testing significance of regression parameters.

@@

Table \ref{tbl:compare} illustrates the lineup protocol in relation to conventional hypothesis testing. Both start from the same place, the same null hypothesis. The conventional test statistic is the $t$-statistic, where the parameter estimate is divided by its standard error. In the lineup protocol, the test statistic is a plot of the data, here side-by-side boxplots are used, because the variable of interest is categorical and takes just two values. In conventional hypothesis testing the value of the test statistic is compared with all possible values of the sampling distribution, the distribution of the statistic if the null hypothesis is true. If it is extreme on this scale then the null hypothesis is rejected. In contrast in visual inference, the plot of the data is compared with plots of a finite number of samples drawn from the null distribution. If the actual data plot is selected as the most different, then this results in rejection of the null hypothesis.

\begin{table*}[hbtp]
\caption{Comparison of visual inference with existing inference technique.  }
\centering 
\begin{tabular}{llll} 
\hline
  & Conventional Inference &  Lineup Protocol \\ %[0.5ex] % inserts table %heading 
\hline
  Hypothesis & $H_0: \beta=0$ vs $H_1: \beta > 0$& $H_0: \beta=0$ vs $H_1: \beta > 0$\\
 & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} \\

@@

%This section outlines the concepts of visual inference in comparison to the procedures of conventional statistical inference. Table \ref{tbl:compare} (derived from \citet{buja:2009}) gives a summarized overview of this comparison.

In general, we define $\theta$ to be a population parameter of interest, with $\theta \in \Theta$, the parameter space. Any null hypothesis $H_0$ then partitions the parameter space into $\Theta_0$ and $\Theta_0^c$, with $H_0: \theta \in \Theta_0$ versus $H_1: \theta \in \Theta_0^c$. A test statistic, $T(y)$, is a function that maps the sample into a numerical summary, that can be used to test the null hypothesis. The hypothesis test maps the test statistic into \{0, 1\}, based on whether $T(y)$ falls into the acceptance region, or the rejection region, respectively. $T(y)$ is assessed relative to null values of this statistic $T(y_0)$ which would be possible values of $T$ if $\theta \in \Theta$.  %Similarly, for any test statistic, we have a rejection region $R$ and, complementary to that, $R^c$,  that partitions the space of the test statistic, which is usually a subspace of the real numbers. 
For visual inference, unlike in the conventional hypothesis test, the statistic is not a single value, but a graphical representation of the data chosen to display the strength of the parameter of interest, $\theta$. When the alternative hypothesis is true, it is expected that the plot of the actual data, the test statistic, will have visible feature(s) consistent with $\theta \in \Theta_0^c$, and that visual artifacts will not distinguish the test statistic as different when $H_1$ is not true. We will call a plot with this property a {\it visual statistic} for $\theta$. More formally,

@@

Power is an important consideration in deciding which test to use for solving a problem. %In conventional testing, new test procedures are examined for their power under a variety of scenarios. 
For visual inference, this corresponds to the choice of plot to use. Analysts typically have a choice of plots to make, and a myriad of possible options such as reference grids, for any particular purpose. This is akin to different choices of statistics in conventional hypothesis testing, for example, mean, median, or trimmed mean. One is typically better than another. For two different visual test statistics of the same actual data, one is considered to be better, if $T(y)$ is more easily distinguishable to the observer. Power is typically used to measure this characteristic of a test. %Power is the probability of rejecting $H_0$ when it is not true.  We can assess power therefore both empirically based on experimental data and through theoretical considerations.

@@

The following properties are only derived for the situation of a  fully parameterized  simulation study. This allows us to directly compare lineup tests against their conventional counterparts and also identify  properties relevant for a quality assessment of  lineups, which are typically used  in non-parametric situations.

@@

%Now suppose $F_{|t|}(.;\delta)$ denotes the distribution function of an absolute value of $t$, the conventional test statistic.
%, with non-centrality parameter $\delta$. 
Then the distribution function of the $p$-value $p_D$ under $H_0$ is uniform, since:  
\begin{eqnarray}\label{dist_p}
F{p_D}(p) &=& Pr(p_D \le p)=1-Pr(1-p_D \le 1-p) \nonumber \\
  &=& 1-Pr\left(F_{|t|}^{-1}(1-p_D) \le F_{|t|}^{-1}(1-p) \right) \nonumber \\
  &=& 1-Pr\left(|t_{obs}| \le F_{|t|}^{-1}(1-p)\right) \nonumber \\
  &=&%\left\{ \begin{array}{ll}
          1-F_{|t|}\left( F_{|t|}^{-1}(1-p)\right)=p \mbox{ ; under $H_0$} 
 %         1-F_{|t|}( F_{|t|}^{-1}(1-p); \delta) &\mbox{ ; under $H_1$} 
%       \end{array} \right.     
\end{eqnarray}

@@

In this setting, there are many established graphics that are used to evaluate and diagnose the fit of a regression model (e.g. \citet{cook:99}). Table \ref{tbl:stat_multiple} lists several common hypotheses related to the regression setting, and the visual test statistic that might be used for testing these hypotheses. These plots should be familiar, because they are all in common use. For example, to examine the effect of variable $X_j$ on $Y$, we would plot $Y$ against $X_j$ (cases 1-4). To assess whether the linear assumption is appropriate we would make a plot of residuals against fitted values (case 5). For the purposes of comparing visual against conventional inference, we focus on cases 2 and 3, with a continuous and catgorical explanatory variable respectively.

@@

% Di: This is specific for the regression model
%Consider the hypothesis test $H_0: \beta_k=0$ against $H_1: \beta_k \ne 0$ in model \ref{multi}.
%%Now consider estimating the power of the visual test. 
%%Suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$. For the regression slope parameter $\beta$ suppose the associated test statistic in conventional test be $t_{obs}$ with $p$-value $p_D$. By definition we have 
%%
%%$$p_D=Pr(|t| \ge t_{obs}| H_0)=1-F_{|t|}(t_{obs}) \Rightarrow |t_{obs}|=F_{|t|}^{-1}(1-p_D)$$
%%
%%\red{HH: the following paragraph should probably move up, but I'm not sure how to change it appropriately. Mahbub, could you take care of that?}
%%Now suppose $F_{|t|}(.;\delta)$ denotes the distribution function of an absolute value of $t$ distribution with non-centrality parameter $\delta$. Thus the distribution function of $p-value$ be  
%%\begin{eqnarray}\label{dist_p}
%%F{p_D}(p) &=& Pr(p_D \le p)=1-Pr(1-p_D \le 1-p) \nonumber \\
%%  &=& 1-Pr(F_{|t|}^{-1}(1-p_D) \le F_{|t|}^{-1}(1-p) ) \nonumber \\
%%  &=& 1-Pr(|t_{obs}| \le F_{|t|}^{-1}(1-p)) \nonumber \\
%%  &=&\left\{ \begin{array}{ll}
%%          1-F_{|t|}( F_{|t|}^{-1}(1-p))=p &\mbox{ ; under $H_0$} \\
%%          1-F_{|t|}( F_{|t|}^{-1}(1-p); \delta) &\mbox{ ; under $H_1$} 
%%       \end{array} \right.     
%%\end{eqnarray}
%%
%%Thus the density of $p_D$ is Uniform(0,1)  under $H_0$. As noted by \cite{Ruppert:2007}, under $H_1$ the density $$f_{p_D}(p_D; \delta)= \frac{f_{|t|}(F_{|t|}^{-1}(1-p_D);\delta)}{f_{|t|}(F_{|t|}^{-1}(1-p_D))}$$ derived from equation \eqref{dist_p}  is a right skewed distribution.  
%In a lineup plot we simulate $m-1$ residual data sets from null model where each of these $m-1$ null data sets produces a corresponding $p$-value $p_{0,i}$ and $p_{0,i} \sim \text{Uniform}(0,1)$ for $i = 1, ..., m-1$. Suppose  $p_0=\min_i(p_{0,i})$. Thus $p_0 \sim \text{Beta}(1,m-1)$.

The conventional test in this scenario uses $T= \hat{\beta_k}/ se(\hat{\beta_k})$ and rejects the null hypothesis if $T$ is extreme on the scale of a $t$ distribution with $n-p$ degrees of freedom.  It forms the benchmark upon which we evaluate the visual test. To calculate what we might expect for the power of the visual test, under perfect conditions, first assume that the observer is able to pick the plot with the smallest $p$-value from a lineup plot.  This leads to the decision to reject $H_0$ when $p_{D} < p_0$, where $p_{D}$ is the conventional $p$-value as details given in Lemma \ref{lemma}. Thus the expected power of the visual test in this scenario is

@@

Three experiments were conducted to evaluate the effectiveness of the lineup protocol relative to the equivalent test statistic used in the regression scenario. The first two are conventional settings where we would not expect the lineup protocol to do better than the conventional test, but we hope it to perform favorably. The third experiment is a scenario where assumptions required for the conventional test are violated, and we would expect the lineup protocol to outperform the conventional test. (Data and lineups used in the experiments are available in the supplementary material.)

@@

%\green{Did you in fact bin like this???}\red{For each parameter combination a set of 1000 actual data are generated and their associated p-values were computed. This produces an empirical distribution of p-value under alternative hypothesis. Finally, three data sets that have 16th, 50th and 84th quantile of the p-values were taken. Similar procedure was used to generate 19 null plots based on the empirical distribution of p-value under null hypothesis}. For added control, to ensure a signal in the simulated actual data a blocking structure was used to filter data sets. A 1000 sets were generated for each parameter combination and the conventional $t$-statistic and $p$-value associated with $H_0: \beta_2=0$ were calculated. The 3 replicates were drawn from each of three blocks of $p$-values: (0.0-$q_{33}$), ($q_{33}$-$q_{66}$), ($q_{66}$-1) where $q_i$ is the $i$th percentile in the distribution of the $p$-values. Additional control was applied to the 19 null plots. Because the distribution of these $p$-values should follow a Uniform(0,1) distribution, data sets were binned on this range by $p$-value, and a data set was randomly selected from each bin.

@@

In the real setting, where visual inference is to be useful, there will be no conventional tests $p$-values. Assessing the strength of perceived structure is a critical component of visual inference. In experiments 1 and 2, there is a $p$-value associated with the actual data plot in each lineup. As the $p$-value increases the proportion of correct responses falls (Figure \ref{fig:pval_pcorrect}), which is evidence of direct association between proportion of correct responses and conventional test $p$-values. When $p$-value is larger than 0.15, it is very uncommon for subjects to correctly identify the actual data plot in the lineup.

@@

%Note that $\hat{p}_D$ is bounded within $(0, 1/(m-1))$. This estimate has to be interpreted carefully, since the choice of a large lineup size, $m$, will make the value look `significant' in a conventional sense. Values close to the upper bound of $1/(m-1)$ should therefore be interpreted as `at least as high', rather than their actual numeric value.
%\hh{For $p$-values below $1/(m-1)$ visual $p$-values are very similar to actual $p$-values, the correlation for those values is: 0.36, 0.79, and -0.48 based on $n_1 = 44, n_2 = 45$, and $n_3= 26$ lineups, respectively.  }  \blue{This is not happening as we see in figure \ref{fig:pval_plot_signal}. But it is true for experiment 2, right?. }
%is equivalent to conventional test $p$-value when $p$-value is less than or equal to $1/(m-1)$.  
%and beyond that the plot signal strength levels off on a point bigger than 0.05. Thus if plot signal strength is less than 0.05, we may decide to reject the null hypothesis.

@@

The purpose of this paper has been to examine the effectiveness of visual inference methods in direct comparison to existing inference methods. We need to be clear that this is not the purpose of visual inference generally: visual methods should not be seen as competitors to conventional inference.  The purpose here, is to  establish properties and  efficacy of visual testing procedures in order to use them in situations where conventional tests cannot be used. For this experiment the effect of $\beta_2$ was examined using side-by-side boxplots. Future experiments will be conducted to compare other regression parameters as described in Table \ref{tbl:stat_multiple} and assess sensitivity of power to modeling conditions.

@@

\begin {abstract} 
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. Until recently there were no formal visual methods in place for determining statistical significance of findings. This changed, when Buja et al.(2009) conceptually introduced two protocols for formal tests of visual findings. In this paper we take this a step further by comparing the lineup protocol (Buja et al.2009) against conventional statistical testing  of the significance of regression model parameters. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol provides results equivalent to the conventional test and for some scenarios yields better power than the conventional test.
\end {abstract}
